---
title: "Lecture 5 - Linear Regression"
author: "Ryan Gallagher"
date: "`r Sys.Date()`"
output: html_document
---

# Boston Dataset Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ISLR2)
library(broom)
```

## Objective

\<What are we analyzing? Why? What insight can we gain from this?\>

## Data Understanding & Preperation

\<What kinds of variables do we have? What kind of questions can we answer with this data?\>

### Data Loading

\<What does the summary say about this data?\>

```{r data prep}
data(Boston)
Boston = as_tibble(Boston) #Tidyverse dataframe

glimpse(Boston)
summary(Boston)

missing_values = Boston %>%
  summarise(across(everything(), ~ sum(is.na(.))))
print(missing_values)
```

### Train-Test Split

\<This is a machine learning technique. How does it aid our analysis?\>

```{r train-test split}
set.seed(123)  # For reproducibility
Boston_split = Boston %>%
  mutate(id = row_number()) %>%
  sample_frac(0.75)

Boston = Boston %>% mutate(id = row_number())

train_data = Boston_split
test_data = anti_join(Boston, Boston_split, by = "id")  
```

### Exploratory Data Analysis

\<What figures did you build? What information do they convey?\>

```{r figures}
ggplot(Boston, aes(x = medv)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Median Home Values", x = "Median Value ($1000s)", y = "Count")

ggplot(Boston, aes(x = lstat, y = medv)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Scatter Plot: LSTAT vs MEDV", x = "% Lower Status Population", y = "Median Home Value ($1000s)")
```

## Model Implementation & Explanation

\<What model are we using? Why does this/these model(s) apply to this data? What are the pros & cons of this type of model? What more can you explain about this model implementation?\>

### Perform Simple Linear Regression on Training Data

\<What is this function? How does it work? What model evaluations do we get from the summary?\>

```{r lm on training}
lm_fit = lm(medv ~ lstat, data = train_data)
summary(lm_fit)
```

### Apply Model to Test Data

\<What does the test mse say about this model?\>

```{r test data}
train_mse = mean((train_data$medv - predict(lm_fit, train_data))^2)
test_mse = mean((test_data$medv - predict(lm_fit, test_data))^2)

print(paste("Training MSE:", round(train_mse, 2)))
print(paste("Test MSE:", round(test_mse, 2)))
```

### Simple Linear Regression Results & Interpretation

\<Overall, how good is this fit? What does it say about the data, and the question being asked?\>

### Perform Multiple Linear Regression on Training Data

\<What question does this model answer?\>

```{r multiple regression training}
lm_fit2 = lm(medv ~ lstat + age, data = train_data)
summary(lm_fit2)
```

### Apply Model to Test Data

\<What does the test mse say about our fit?\>

```{r multiple regression test}
train_mse2 = mean((train_data$medv - predict(lm_fit2, train_data))^2)
test_mse2 = mean((test_data$medv - predict(lm_fit2, test_data))^2)

print(paste("Training MSE:", round(train_mse2, 2)))
print(paste("Test MSE:", round(test_mse2, 2)))
```

### Multiple Linear Regression Results & Interpretation

# NHANES Dataset Analysis

## Objective

**Please predict BMI using predictors Age, SmokeNow, and PhysActive**

\<What are we analyzing? Why? What insight can we gain from this?\>

## Data Understanding & Preperation

\<What kinds of variables do we have? What kind of questions can we answer with this data?\>

### Data Loading

\<What does the summary say about this data?\>

```{r load NHANES}
library(NHANES)
data(NHANES)
SMOKERS <- NHANES %>%
  select(BMI, Age, SmokeNow, PhysActive) %>%
  drop_na() %>%
  filter(Age >= 18 & Age <= 70) %>%
  mutate(
    SmokeNow = as.factor(SmokeNow),
    PhysActive = as.factor(PhysActive)
  )


```

### Train-Test Split

\<This is a machine learning technique. How does it aid our analysis?\>

### Exploratory Data Analysis

\<What figures did you build? What information do they convey?\>

## Model Implementation & Explanation

\<What model are you using? Why does this/these model(s) apply to this data? What are the pros & cons of this type of model? What more can you explain about this model implementation?\>

## Results and Interpretation

\< What is our interpretation of the model you'be built? How have you answered the question posed? \>
