---
title: "Linear Regression in R"
author: "Ryan Gallagher"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

#install.packages("ISLR2")
library(ISLR2)
```

# Boston Dataset Analysis

### Objective

How can we predict the Median Value of Owner-Cccupied Homes using the Lower Status of Population metric.

\<What are we analyzing? Why? What insight can we gain from this analysis?\>

### Data Understand & Preperation

\<What kind of variables do we have? What potential questions could we answer further with this data?\>

### Data Exploration

\<What do the summaries say about our data?\>

```{r load.data}
data(Boston)
glimpse(Boston)

summary(Boston)
```

```{r missing values}
missing_values = Boston %>%
  summarise(across(everything(), ~ sum(is.na(.))))
print(missing_values)
```

### Train-Test Split

\<How does this technique aid our analysis, especially given new data?\>

```{r train-test}
set.seed(123) # For reproducibility
Boston_split = Boston %>%
  mutate(id = row_number()) %>%
  sample_frac(0.75)

Boston = Boston %>% mutate(id = row_number())

train_data = Boston_split
test_data = anti_join(Boston, Boston_split, by = "id") #Remaining 25%

```

### Exploratory Data Analysis

\<What figures did we build? Why? What information do they convey? How is it important to the analysis?\>

```{r histogram for medv}
ggplot(Boston, aes(x = medv)) +
  geom_histogram(fill = "steelblue", binwidth = 2, color = "white") + 
  labs(title = "Distribution of Median Home Values",
       x = "Median Value ($1000s)",
       y = "Count")
```

```{r LSTAT vs MEDV Scatterplot}
ggplot(Boston, aes(x = lstat, y=medv)) +
  geom_point(alpha = 0.6, color = 'blue') +
  labs(title = "Scatterplot: LSTAT vs. MEDV",
       x = "% Lower Status Population",
       y = "Median Home Value ($1000s)")
```

### Model Implementation & Explanation

\<What model are we using? Why does this/these model(s) apply to the data? What are the pros & cons of this type of model? \>

#### Perform Simple Linear Regression on Training Data

\<Describe the function & model fit. Maybe talk about the evaluation metrics?\>

```{r linear regression}
lm.fit = lm(medv ~ lstat, data = train_data)
summary(lm.fit)
```

Could build a scatter plot with this regression line onto it.

### Apply Model to Test Data

\<Could interpret the Test MSE\>

```{r apply model to test_data}
train_mse = mean((train_data$medv - predict(lm.fit, train_data))^2)
test_mse = mean((test_data$medv - predict(lm.fit, test_data))^2)

print(paste("Training MSE:", round(train_mse, 2)))
print(paste("Test MSE:", round(test_mse, 2)))
```

### Simple Linear Regression Results & Interpretation

\<Overall, how good is this fit? What does it say about the data and the question being asked?\>

### Perform Multiple Linear Regression on Training Data

\<What question does this model answer?\>

```{r}
lm.multiple.fit = lm(medv ~ lstat + age, data=train_data)
summary(lm.multiple.fit)
```

### Apply the Model to Test Data

```{r}
train_mse = mean((train_data$medv - predict(lm.multiple.fit, train_data))^2)
test_mse = mean((test_data$medv - predict(lm.multiple.fit, test_data))^2)

print(paste("Training MSE:", round(train_mse, 2)))
print(paste("Test MSE:", round(test_mse, 2)))
```

### Multiple Linear Regression Results & Interpretation

\<Interpretation\>

# NHANES Data Analysis

## Objective

**Please predict BMI using Age, SmokeNow, PhysActive for individuals between the ages of 18 and 70**

## Data Understanding & Preperation

### Data Loading

```{r}
#install.packages("NHANES")
library(NHANES)
data(NHANES)

SMOKERS = NHANES %>% 
  select(BMI, Age, SmokeNow, PhysActive) %>%
  filter(Age >= 18 & Age <= 70)
```
